Books and Babies: An Analysis of the Relationship between Female Literacy Rate and Fertility Rate in South Asia (2000-2020)

### 2024-04-19 paper/paper.qmd
update references.bib and paper.qmd

According to the World Bank, the South Asian region comprises Afghanistan, Bangladesh, Bhutan, India, Maldives, Nepal, Pakistan, and Sri Lanka. It is the most populated region in the world with countries like India, Pakistan and Bangladesh which ranks among the top 10 in in terms of population size. Majority of these countries are also considered developing nations and play a major role in the world economy. In today’s globalized world where population dynamics significantly influence a country's course it is necessary to analyze gender statistics indicators such as fertility rate and female literacy rate and how they relate to each other. By doing so we can gain valuable insights which can help make policy decisions on female education and empowerment that inturn help manage population growth and promote sustainable development.

@@

The data used in this paper was obtained from the World Bank's Data Bank which is publicly available on the Data Bank website [@The_World_Bank_2024a]. A considerable amount of the data in the World Bank's Data Bank is gathered and compiled by the World Bank itself, along with information from other reliable public and private sector sources, international organizations, national statistical offices, and specialized agencies of the United Nations (UN). The World Bank then compiles and examines these data to find patterns and provide guidance in policy matters.

The two datasets that are used in this paper are fertility rate (total births per woman)[@The_World_Bank_2024c] and literacy rate, adult females (% of females ages 15 and above) [@The_World_Bank_2024b]. Both female literacy rate and fertility rate data was available for a total of 266 regions including individual countries and group of countries (for example - South Asia, Middle East etc.). While the literacy rate data was available from year 1960 to 2022, the fertility rate data consisted of data from 1960 to 2021. Not every country had data available for each year. The region of interest for this study is South Asia and the time period selected was 2000-2020. Both fertility rate and female literacy rate data was available for South Asia for the specified time period. The two indicators were merged to form a new data set with 21 rows that gives both fertility rate and female literacy rate in South Asia for any given year between 2000-2020.Total fertility rate represents the number of children that would be born to a woman if she were to live to the end of her childbearing years and bear children in accordance with age-specific fertility rates of the specified year. Female Adult literacy rate is the percentage of females ages 15 and above who can both read and write with understanding a short simple statement about their everyday life.

@@

library(tidyverse)
library(here)
library(readr)
library(dplyr)
library(ggplot2)
library(modelsummary)

analysis_data <- read.csv(here::here("data/analysis_data/cleaned_data_south_asia_2000_2020.csv"))

@@

kable(analysis_data)

```

@@

```{r}
#| label: fig-3
#| fig-cap: "The model shows the relationship between Female Literacy Rate and Fertility Rate in South Asia (2000-2020)"
#| echo: false
#| warning: false
library(ggplot2)

analysis_data %>%
  ggplot(aes(x = `female_literacy_rate`, y = `fertility_rate_total`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "solid") +
  labs(x = "Literacy Rate, Adult Females (% of females ages 15 and above)", y = "Fertility Rate, Total (births per woman)") +
  theme_classic()
```

@@

As analyzed in @fig-1,  the female literacy rate in South Asia is on the rise over the last two decades with a literacy rate of approximately 44.9 percent in the year 2000 to approximately 64.6 percent in 2020. This upward trend is not confined to the past two decades only. South Asia has made a tremendous progress in female literacy rate which was only 21.27 percent in 1975[@The_World_Bank_2024b].

@@

As analyzed in @fig-2,  the fertility rate in South Asia is on the decline over the last two decades with a fertility rate of approximately 3.57 births per woman in the year 2000 to approximately 2.26 births per woman in 2020. This downward trend is not confined to the past two decades only. Fertility rate in South Asia has long been on a decline. The South Asian fertility rate was as high as 6.07 births per woman in 1960[@The_World_Bank_2024c].

>>>>

According to the World Bank, the South Asian region comprises Afghanistan, Bangladesh, Bhutan, India, Maldives, Nepal, Pakistan, and Sri Lanka. It is the most populated region in the world with countries like India, Pakistan and Bangladesh which ranks among the top 10 in in terms of population size. Majority of these countries are also considered developing nations and play a major role in the world economy. In today’s globalized world where population dynamics significantly influence a country's course it is necessary to analyze gender statistics indicators such as fertility rate and female literacy rate and how they relate to each other. By doing so we can gain valuable insights which can help make policy decisions on female education and empowerment that in-turn help manage population growth and promote sustainable development.

@@

The data used in this paper was obtained from the World Bank's Data Bank which is publicly available on the Data Bank website [@The_World_Bank]. A considerable amount of the data in the World Bank's Data Bank is gathered and compiled by the World Bank itself, along with information from other reliable public and private sector sources, international organizations, national statistical offices, and specialized agencies of the United Nations (UN). The World Bank then compiles and examines these data to find patterns and provide guidance in policy matters.

The two data sets that are used in this paper are fertility rate (total births per woman)[@The_World_Bank_Fertility] and literacy rate, adult females (% of females ages 15 and above) [@The_World_Bank_Literacy]. Both female literacy rate and fertility rate data was available for a total of 266 regions including individual countries and group of countries (for example - South Asia, Middle East etc.). While the literacy rate data was available from year 1960 to 2022, the fertility rate data consisted of data from 1960 to 2021. Not every country had data available for each year. The region of interest for this study is South Asia and the time period selected was 2000-2020. Both fertility rate and female literacy rate data was available for South Asia for the specified time period. The two indicators were merged to form a new data set with 21 rows that gives both fertility rate and female literacy rate in South Asia for any given year between 2000-2020.Total fertility rate represents the number of children that would be born to a woman if she were to live to the end of her childbearing years and bear children in accordance with age-specific fertility rates of the specified year. Female Adult literacy rate is the percentage of females ages 15 and above who can both read and write with understanding a short simple statement about their everyday life.

@@

# Load necessary libraries
library(tidyverse)
library(here)
library(readr)
library(dplyr)
library(ggplot2)
library(modelsummary)

@@

kable(analysis_data)
```

@@

```{r}
#| label: fig-3
#| fig-cap: "The model shows the relationship between Female Literacy Rate and Fertility Rate in South Asia (2000-2020)"
#| echo: false
#| warning: false
# Load necessary libraries
library(ggplot2)

# Create ggplot
analysis_data %>%
  ggplot(aes(x = `female_literacy_rate`, y = `fertility_rate_total`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "solid") +
  labs(x = "Literacy Rate, Adult Females (% of females ages 15 and above)", y = "Fertility Rate, Total (births per woman)") +
  theme_classic()
```

@@

As analyzed in @fig-1,  the female literacy rate in South Asia is on the rise over the last two decades with a literacy rate of approximately 44.9 percent in the year 2000 to approximately 64.6 percent in 2020. This upward trend is not confined to the past two decades only. South Asia has made a tremendous progress in female literacy rate which was only 21.27 percent in 1975[@The_World_Bank_Literacy].

@@

As analyzed in @fig-2,  the fertility rate in South Asia is on the decline over the last two decades with a fertility rate of approximately 3.57 births per woman in the year 2000 to approximately 2.26 births per woman in 2020. This downward trend is not confined to the past two decades only. Fertility rate in South Asia has long been on a decline. The South Asian fertility rate was as high as 6.07 births per woman in 1960[@The_World_Bank_Fertility].
